Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2025-05-03 article1.qmd
Improvements to tables in pdf render.
Mac work session. Appendix finished. All refs to OSF repos completed.

---
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/creating-frames/"
  fig-dpi: 600

---

@@

# Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia
Joana Belmiro^1^, Xavier Terradas^2^, João Cascalheira^1^

@@

The Algarve is the southernmost region of Portugal, framed north by the Alentejo region and east by Spain. To the west and south, it is bordered by the Atlantic ocean. It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (\@fig-frames-geomap). On the north sector of the Algarve, the Serra Algarvia is characterized by a mountainous range with a dense hydrographic network, which separates the Algarve from Alentejo. On the south sector, the Litoral is characterized by a flatter, long strip of land, that extends through all of the coastal strip of the Algarve. The Barrocal is nested between the other sub-regions and has a more moderate relief, characterized by carbonated Jurassic formations and important sub-terranean water circulation [@gago_aquifero_2007].

@@

A two-step approach was applied to characterize the geological samples. These include 59 geological samples (Table S2), currently located at ICArEHB’s laboratories (University of Algarve). No permits were required for the described study, which complied with all relevant regulations. The samples were analyzed macroscopically following a pre-established dataset. The variables were defined based on specialized literature [@bressyCaracterisationGestionSilex2002; @crandellMacroscopicAnalysisCharacterisation2005; @luedtkeArchaeologistGuideChert1992], and the dataset with the sample characterizations as well as the variable descriptions can be found in the SOM (Tables S2 and S3). A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content. Despite several caveats, especially related to the subjectivity and lack of quantitative variables [@bustilloMacroscopicClassificationFlint2009], a macroscopic approach is currently still frequently used in chert raw material studies. For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages. Macroscopic analyses have the advantage of being less costly and easy to apply. Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential for comparative studies between geological samples and archaeological assemblages and obtaining preliminary results. By then combining the macroscopic analysis with a petrographic analysis, other studies have shown that the subjective component of this approach can be minimized [@bustilloMacroscopicClassificationFlint2009]. The second phase of the study focused on the petrographic analysis of the geological samples. Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability. In total, 30 thin sections were produced (\@tbl-frames-petrolist), divided into three groups: 1) 20 thin sections of geological samples from different outcrops of the Lower Jurassic and Upper Jurassic chert-bearing formations within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops of the Malhão and Jordana chert-bearing formations, from the eastern section of the Algarve; 3) 1 thin section of a geological sample recovered from previous works, which was not identified during our survey. Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced. To compare with the thin sections from this study, other thin sections from previous studies of Jurassic outcrops from western Algarve [@ribeiro_evolucao_2005] were also consulted. All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in S3 Table).

@@

petrolist <- read_csv(here::here("data/raw_data_frames/petro_sample.csv"))
knitr::kable(petrolist)

@@

On the westernmost areas of the Algarve, there are mainly cherts from two different formations: Carixian formations (Lower Jurassic) and Kimmeridgian formations (Upper Jurassic, @fig-frames-sourcesmap). The latter can be found in primary deposition in a single known outcrop - Praia da Mareta (MAR) – or nearby, in secondary deposition settings. Lower Jurassic outcrops are more common and, for that reason, have been more studied [@ribeiro_evolucao_2005]. These outcrops are heterogeneous, showing different geological characteristics and chert colors (\@fig-frames-ljmacro). The Lower Jurassic cherts can be grouped into three main macroscopic types based on color (individual Munsell Color Chart codes can be found in the macroscopic description analysis table) and presence of fossil content: 1) multicolored, yellow, red, light gray or purple type (MC, @fig-frames-ljmacro b and @fig-frames-ljmacro d); 2) single grey/brown type (SGB, @fig-frames-ljmacro e and @fig-frames-ljmacro f); 3) multicolored, yellow, red, light gray or purple with fossils type (MCF, @fig-frames-ljmacro a and @fig-frames-ljmacro c). The first two types are present in all outcrops. They are mainly characterized by dull to medium luster and opaque translucency, although some samples were sub-translucent. The feel ranges between smooth and semi-smooth, although many of the cherts from the Belixe outcrop are distinctly rough to the touch. In the MC cherts, fossil content is present but visible only as white, red, or yellow speckling. The SGB cherts show little fossil content, barely visible with the stereomicroscope. The MCF show a large quantity of larger fossils (\~1000 µm), which are easily seen by the naked eye and can be identified under the microscope.

@@

Petrographically, the Lower Jurassic cherts of western Algarve are composed mainly of microcrystalline quartz, with textures that range mostly from wackestone to packstone (\@fig-frames-ljwest). Dolomite is also present although in frequencies inferior to 10%. Macrocrystalline quartz and chalcedony occur in small frequencies (\~5%), frequently replacing bioclasts. The presence of mica is uncommon and always below 1%. Allochems present are mostly iron oxides (ranging from uncommon to very frequent) and the presence of peloids is rare. In more than 50% of the samples, no fossils can be identified, as all fossils, albeit common to very frequent, are poorly preserved, and without any identifiable morphology. Whenever identifiable, fossils present in the sample are Echinoderms (\@fig-frames-ljwest b and \@fig-frames-ljwest c), Radiolarians, Sponge spicules, and bivalve shells (\@fig-frames-ljwest e and @fig-frames-ljwest f). In all samples porosity ranges from 1-5% (vuggy type).

@@

Contrasting with the diversity and quantity of the Lower Jurassic outcrops, there is only one identified outcrop for Upper Jurassic cherts in western Algarve, located at Praia da Mareta (MAR) and abundant, or in a secondary deposition at Ponta da Atalaia (PtA). The Upper Jurassic cherts are very similar to the Lower Jurassic, with dull to medium luster and grey/purple colors (\@fig-frames-ujwest a). The translucency ranges from opaque to areas where the chert is translucent. This translucency may be a significant difference to distinguish between outcrops. Petrographically, the cherts are also similar to the Lower Jurassic ones. The only identifiable difference is the presence of calcispheres. All samples from the MAR and PtA outcrops seen under the petrographic microscope showed the presence of abundant calcispheres (\@fig-frames-ujwest b and @fig-frames-ujwest c), which is not always apparent with the stereomicroscope. Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic (\@fig-frames-ujwest e and @fig-frames-ujwest f), which were uncommon and scattered at the top of the cliffs by the beach.

@@

On the eastern part of the Algarve, chert-bearing known formations are from the Middle and Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively. The Malhão formation chert (Middle Jurassic) was identified in three outcrops: 1) Casal da Colina (CdC); 2) Oliveiras (OLV); and 3) Malhão (MALH). Whenever in a primary outcrop, this chert was homogeneous. Secondary deposits were also identified—Casal da Colina (CdC), Guilhim (GUI), Caliços (CAL) and Oliveiras (OLV) and were located in recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations (in these cases, it was not possible to confirm the outcrop location). In the Malhão formation outcrops, the nodule frequency varied from common to abundant. The nodules are roundish, ranging between 3 to 5 cm in maximum width. In all cases, access to the outcrops was easy. Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits. The Malhão cherts show two differing macroscopic characteristics: pink/reddish/light gray cherts (\@fig-frames-mjplate b) and grey cherts (\@fig-frames-mjplate a). In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel. They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white speckling in plain sight. Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified. The petrographic analysis shows that the Malhão formation cherts are characterized by a wackestone texture, and composed of microcrystalline quartz (85-95%). Dolomite is also present (10-5%), as well as chalcedony and macrocrystalline quartz (\<5%) frequently replacing fossils. Identified allochems are oxide patina, ranging from very frequent to uncommon. A high variety of identifiable fossils (although all are poorly preserved (\@fig-frames-mjplate) were also identified. These fossils are Sponge spicules (\@fig-frames-mjplate e and @fig-frames-mjplate f), Radiolarians, Ostracods (\@fig-frames-mjplate c and @fig-frames-mjplate d), Echinoderms, Calcispheres, and possibly Tentaculites. Porosity in the samples occurs in small frequencies (\<5%), of vuggy type.

@@

The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (\@fig-frames-sourcesmap), in an outcrop of the same name (JOR). Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock. No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape. Smaller nodules broken from the parent rock were identified near the primary source in a field. Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock. The cherts show little macroscopic variability between nodule and outcrop. They are grey/brown (with visible yellow inclusions) (\@fig-frames-ujplate a and @fig-frames-ujplate b). Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas. Some of the nodules also show variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions. The petrographic analysis shows that the cherts range from a wackestone to packstone texture (\@fig-frames-ujplate), which was already seen macroscopically. They are composed mostly of microcrystalline quartz (90-99%), with the presence of fibrous chalcedony (1%) replacing the fossils and dolomite (1%), as well as negligible percentages of other minerals. Present allochems are iron oxides, ranging between very frequent to common. Albeit frequent, fossils are poorly preserved in general, with a few being identifiable: Calcispheres (\@fig-frames-ujplate c and @fig-frames-ujplate d), Bivalve shell (\@fig-frames-ujplate e and @fig-frames-ujplate f), Sponge spicules, Ostracod, Echinoderms, and Gastropod. Porosity is small (\~1%) of vuggy type.

@@

The survey work and analyses of the collected geological samples show that the south of Portugal has a high potential for chert raw material studies. The presence of chert-bearing geological formations throughout the Algarve would provide several possibilities for sourcing and procurement whenever groups moved throughout the territory. This is further important when we consider the geology of this region. The geology of the Algarve itself may have played an important part in how groups procured their raw materials, specifically, their chert, a task that has been identified as essential for hunter-gatherer groups. To the south, communities would only have access to chert-bearing outcrops down to the coast. To the north, the mountain range would not only have provided no chert nodules but may have also hampered the movement of populations, forcing groups to move east and west instead of north or south. This movement may have facilitated the gathering of cherts from different formations within the Algarve, posteriorly then brought into the sites. Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory these groups were sourcing their chert raw materials, and how they were using the territory having in consideration the region’s natural barriers and consequent distribution of resources. Although this topic remains unexplored, this study stands as a further step to tackle these questions in the Algarve, as it may provide the necessary basis for comparative studies with archaeological assemblages. However, tracking these movements and procurement patterns is only possible if the cherts from the different formations and outcrops can be traced back to their sources. This presented itself as the first caveat for this type of study, since for the Algarve, for example, all cherts formed in Jurassic formations in pelagic environments. Despite the similar formation environments, in general, there seem to be relevant differences between the cherts of different formations and periods. This is further relevant given the fact that they are geographically distant. Within formations, however, there are no discernable differences, both at a macroscopic and petrographic level, as these do not seem to be useful to distinguish between outcrops. This is most obvious on the Lower Jurassic formation of western Algarve. The identified chert groups, which varied mostly in color and fossil content, are present in several outcrops from this formation. In this region, the variables which may be better used to understand which outcrops were visited may be the quality (differing levels of dolomitization, presence/absence of fractures or even size of grain) and size of the nodules. The latter, for example, is an important variable in the Praia do Belixe outcrops, which show the largest volumes of rock, even if the chert’s quality is worse than some other available, smaller nodules. Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred in relation to other smaller nodules in possibly closer outcrops in the region. The Upper Jurassic nodules of western Algarve also show larger volumes than those from the Lower Jurassic. Translucency also seems to be a good macroscopic indicator to distinguish between western Algarve Lower and Upper Jurassic cherts, since the latter are characterized as sub-translucent. However, for a reliable distinction between the Lower Jurassic and the Upper Jurassic cherts, petrographic analyses and the identification of calcispheres may be necessary. The differences identified among the cherts of the various formations can be seen both at a macroscopic and petrographic level. Given the formation settings, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic. The use of specific fossils for the identification of the cherts is also difficult since these are often not well preserved enough to allow the identification of species that may connect a group of cherts. The size, frequency, and preservation state of the fossils seem to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas. These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations – West (including the Lower Jurassic and Upper Jurassic formation cherts of western Algarve), Jordana, and Malhão - to be differentiated without the need for thin sections (\@fig-frames-macrocomparison). An exception might be the distinction between the cherts from the West and Malhão - the reddish cherts from Malhão are visually indistinguishable from those from the West with a higher fossil content. The grey cherts of the Malhão Middle Jurassic formation do show a higher concentration of visible round fossils, however the distinction is only possible seen under the stereomicroscope and on a fresh surface, which might hamper the classification of archaeological materials. These distinctions are especially important for archaeological collections, especially those which may be small, with small artifacts, or for the study of older collections to which other (destructive) means of analysis may not be applied.

@@

## Supporting information {.unnumbered}

S1 Table. Field sample/outcrop dataset. Table with recorded data on the outcrops and cherts, during fieldwork and prospections during 2021 and 2022. S2 Table. Macroscopic analysis dataset. Table with the data collected from the macroscopic analyses of all chert samples collected from the fieldwork and prospections during 2021 and 2022. S3 Table. Data dictionaries for macroscopic and petrographic analyses. Description of the used variables (including allowed variables and references) for the macroscopic and petrographic analyses used in the present study.

@@

## Author contributions {.unnumbered}

**Conceptualization**: Joana Belmiro, Xavier Terradas, João Cascalheira. **Funding acquisition**: Joana Belmiro, João Cascalheira. **Investigation**: Joana Belmiro. **Methodology**: Joana Belmiro, Xavier Terradas, João Cascalheira. **Writing – original draft**: Joana Belmiro. **Writing – review & editing**: Xavier Terradas, João Cascalheira.

>>>>

---
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/creating-frames/"
  fig-dpi: 600
---

@@

# Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia {#sec-chapter2}

Joana Belmiro^1^, Xavier Terradas^2^, João Cascalheira^1^

@@

The Algarve is the southernmost region of Portugal, framed north by the Alentejo region and east by Spain. To the west and south, it is bordered by the Atlantic ocean. It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (@fig-frames-geomap). On the north sector of the Algarve, the Serra Algarvia is characterized by a mountainous range with a dense hydrographic network, which separates the Algarve from Alentejo. On the south sector, the Litoral is characterized by a flatter, long strip of land, that extends through all of the coastal strip of the Algarve. The Barrocal is nested between the other sub-regions and has a more moderate relief, characterized by carbonated Jurassic formations and important sub-terranean water circulation [@gago_aquifero_2007].

@@

A two-step approach was applied to characterize the geological samples. These include 59 geological samples (Table S2), currently located at ICArEHB’s laboratories (University of Algarve). No permits were required for the described study, which complied with all relevant regulations. The samples were analyzed macroscopically following a pre-established dataset. The variables were defined based on specialized literature [@bressyCaracterisationGestionSilex2002; @crandellMacroscopicAnalysisCharacterisation2005; @luedtkeArchaeologistGuideChert1992], and the dataset with the sample characterizations as well as the variable descriptions can be found in the SOM (Tables S2 and S3). A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content. Despite several caveats, especially related to the subjectivity and lack of quantitative variables [@bustilloMacroscopicClassificationFlint2009], a macroscopic approach is currently still frequently used in chert raw material studies. For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages. Macroscopic analyses have the advantage of being less costly and easy to apply. Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential for comparative studies between geological samples and archaeological assemblages and obtaining preliminary results. By then combining the macroscopic analysis with a petrographic analysis, other studies have shown that the subjective component of this approach can be minimized [@bustilloMacroscopicClassificationFlint2009]. The second phase of the study focused on the petrographic analysis of the geological samples. Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability. In total, 30 thin sections were produced (@tbl-frames-petrolist), divided into three groups: 1) 20 thin sections of geological samples from different outcrops of the Lower Jurassic and Upper Jurassic chert-bearing formations within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops of the Malhão and Jordana chert-bearing formations, from the eastern section of the Algarve; 3) 1 thin section of a geological sample recovered from previous works, which was not identified during our survey. Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced. To compare with the thin sections from this study, other thin sections from previous studies of Jurassic outcrops from western Algarve [@ribeiro_evolucao_2005] were also consulted. All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in S3 Table).

@@

petrolist <- read_csv(here::here("data/raw_data_frames/petro_sample.csv"))
knitr::kable(petrolist) |> 
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"))

@@

On the westernmost areas of the Algarve, there are mainly cherts from two different formations: Carixian formations (Lower Jurassic) and Kimmeridgian formations (Upper Jurassic, @fig-frames-sourcesmap). The latter can be found in primary deposition in a single known outcrop - Praia da Mareta (MAR) – or nearby, in secondary deposition settings. Lower Jurassic outcrops are more common and, for that reason, have been more studied [@ribeiro_evolucao_2005]. These outcrops are heterogeneous, showing different geological characteristics and chert colors (@fig-frames-ljmacro). The Lower Jurassic cherts can be grouped into three main macroscopic types based on color (individual Munsell Color Chart codes can be found in the macroscopic description analysis table) and presence of fossil content: 1) multicolored, yellow, red, light gray or purple type (MC, @fig-frames-ljmacro b and @fig-frames-ljmacro d); 2) single grey/brown type (SGB, @fig-frames-ljmacro e and @fig-frames-ljmacro f); 3) multicolored, yellow, red, light gray or purple with fossils type (MCF, @fig-frames-ljmacro a and @fig-frames-ljmacro c). The first two types are present in all outcrops. They are mainly characterized by dull to medium luster and opaque translucency, although some samples were sub-translucent. The feel ranges between smooth and semi-smooth, although many of the cherts from the Belixe outcrop are distinctly rough to the touch. In the MC cherts, fossil content is present but visible only as white, red, or yellow speckling. The SGB cherts show little fossil content, barely visible with the stereomicroscope. The MCF show a large quantity of larger fossils (\~1000 µm), which are easily seen by the naked eye and can be identified under the microscope.

@@

Petrographically, the Lower Jurassic cherts of western Algarve are composed mainly of microcrystalline quartz, with textures that range mostly from wackestone to packstone (@fig-frames-ljwest). Dolomite is also present although in frequencies inferior to 10%. Macrocrystalline quartz and chalcedony occur in small frequencies (\~5%), frequently replacing bioclasts. The presence of mica is uncommon and always below 1%. Allochems present are mostly iron oxides (ranging from uncommon to very frequent) and the presence of peloids is rare. In more than 50% of the samples, no fossils can be identified, as all fossils, albeit common to very frequent, are poorly preserved, and without any identifiable morphology. Whenever identifiable, fossils present in the sample are Echinoderms (@fig-frames-ljwest b and @fig-frames-ljwest c), Radiolarians, Sponge spicules, and bivalve shells (@fig-frames-ljwest e and @fig-frames-ljwest f). In all samples porosity ranges from 1-5% (vuggy type).

@@

Contrasting with the diversity and quantity of the Lower Jurassic outcrops, there is only one identified outcrop for Upper Jurassic cherts in western Algarve, located at Praia da Mareta (MAR) and abundant, or in a secondary deposition at Ponta da Atalaia (PtA). The Upper Jurassic cherts are very similar to the Lower Jurassic, with dull to medium luster and grey/purple colors (@fig-frames-ujwest a). The translucency ranges from opaque to areas where the chert is translucent. This translucency may be a significant difference to distinguish between outcrops. Petrographically, the cherts are also similar to the Lower Jurassic ones. The only identifiable difference is the presence of calcispheres. All samples from the MAR and PtA outcrops seen under the petrographic microscope showed the presence of abundant calcispheres (@fig-frames-ujwest b and @fig-frames-ujwest c), which is not always apparent with the stereomicroscope. Based on the presence of calcispheres, we may also consider the samples recovered at Andorinha (AND) to be Upper Jurassic (@fig-frames-ujwest e and @fig-frames-ujwest f), which were uncommon and scattered at the top of the cliffs by the beach.

@@

On the eastern part of the Algarve, chert-bearing known formations are from the Middle and Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively. The Malhão formation chert (Middle Jurassic) was identified in three outcrops: 1) Casal da Colina (CdC); 2) Oliveiras (OLV); and 3) Malhão (MALH). Whenever in a primary outcrop, this chert was homogeneous. Secondary deposits were also identified—Casal da Colina (CdC), Guilhim (GUI), Caliços (CAL) and Oliveiras (OLV) and were located in recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations (in these cases, it was not possible to confirm the outcrop location). In the Malhão formation outcrops, the nodule frequency varied from common to abundant. The nodules are roundish, ranging between 3 to 5 cm in maximum width. In all cases, access to the outcrops was easy. Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits. The Malhão cherts show two differing macroscopic characteristics: pink/reddish/light gray cherts (@fig-frames-mjplate b) and grey cherts (@fig-frames-mjplate a). In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel. They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white speckling in plain sight. Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified. The petrographic analysis shows that the Malhão formation cherts are characterized by a wackestone texture, and composed of microcrystalline quartz (85-95%). Dolomite is also present (10-5%), as well as chalcedony and macrocrystalline quartz (\<5%) frequently replacing fossils. Identified allochems are oxide patina, ranging from very frequent to uncommon. A high variety of identifiable fossils (although all are poorly preserved (@fig-frames-mjplate) were also identified. These fossils are Sponge spicules (@fig-frames-mjplate e and @fig-frames-mjplate f), Radiolarians, Ostracods (@fig-frames-mjplate c and @fig-frames-mjplate d), Echinoderms, Calcispheres, and possibly Tentaculites. Porosity in the samples occurs in small frequencies (\<5%), of vuggy type.

@@

The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (@fig-frames-sourcesmap), in an outcrop of the same name (JOR). Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock. No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape. Smaller nodules broken from the parent rock were identified near the primary source in a field. Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock. The cherts show little macroscopic variability between nodule and outcrop. They are grey/brown (with visible yellow inclusions) (@fig-frames-ujplate a and @fig-frames-ujplate b). Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas. Some of the nodules also show variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions. The petrographic analysis shows that the cherts range from a wackestone to packstone texture (@fig-frames-ujplate), which was already seen macroscopically. They are composed mostly of microcrystalline quartz (90-99%), with the presence of fibrous chalcedony (1%) replacing the fossils and dolomite (1%), as well as negligible percentages of other minerals. Present allochems are iron oxides, ranging between very frequent to common. Albeit frequent, fossils are poorly preserved in general, with a few being identifiable: Calcispheres (@fig-frames-ujplate c and @fig-frames-ujplate d), Bivalve shell (@fig-frames-ujplate e and @fig-frames-ujplate f), Sponge spicules, Ostracod, Echinoderms, and Gastropod. Porosity is small (\~1%) of vuggy type.

@@

The survey work and analyses of the collected geological samples show that the south of Portugal has a high potential for chert raw material studies. The presence of chert-bearing geological formations throughout the Algarve would provide several possibilities for sourcing and procurement whenever groups moved throughout the territory. This is further important when we consider the geology of this region. The geology of the Algarve itself may have played an important part in how groups procured their raw materials, specifically, their chert, a task that has been identified as essential for hunter-gatherer groups. To the south, communities would only have access to chert-bearing outcrops down to the coast. To the north, the mountain range would not only have provided no chert nodules but may have also hampered the movement of populations, forcing groups to move east and west instead of north or south. This movement may have facilitated the gathering of cherts from different formations within the Algarve, posteriorly then brought into the sites. Especially for Middle and Upper Paleolithic occupations, understanding the sources of chert in the Algarve may provide data about where in the territory these groups were sourcing their chert raw materials, and how they were using the territory having in consideration the region’s natural barriers and consequent distribution of resources. Although this topic remains unexplored, this study stands as a further step to tackle these questions in the Algarve, as it may provide the necessary basis for comparative studies with archaeological assemblages. However, tracking these movements and procurement patterns is only possible if the cherts from the different formations and outcrops can be traced back to their sources. This presented itself as the first caveat for this type of study, since for the Algarve, for example, all cherts formed in Jurassic formations in pelagic environments. Despite the similar formation environments, in general, there seem to be relevant differences between the cherts of different formations and periods. This is further relevant given the fact that they are geographically distant. Within formations, however, there are no discernable differences, both at a macroscopic and petrographic level, as these do not seem to be useful to distinguish between outcrops. This is most obvious on the Lower Jurassic formation of western Algarve. The identified chert groups, which varied mostly in color and fossil content, are present in several outcrops from this formation. In this region, the variables which may be better used to understand which outcrops were visited may be the quality (differing levels of dolomitization, presence/absence of fractures or even size of grain) and size of the nodules. The latter, for example, is an important variable in the Praia do Belixe outcrops, which show the largest volumes of rock, even if the chert’s quality is worse than some other available, smaller nodules. Size may be used in conjunction with other technological data, to understand whether different nodules were being explored differently based on their size, or their procurement was being preferred in relation to other smaller nodules in possibly closer outcrops in the region. The Upper Jurassic nodules of western Algarve also show larger volumes than those from the Lower Jurassic. Translucency also seems to be a good macroscopic indicator to distinguish between western Algarve Lower and Upper Jurassic cherts, since the latter are characterized as sub-translucent. However, for a reliable distinction between the Lower Jurassic and the Upper Jurassic cherts, petrographic analyses and the identification of calcispheres may be necessary. The differences identified among the cherts of the various formations can be seen both at a macroscopic and petrographic level. Given the formation settings, petrographically, all the cherts from the Algarve are fairly homogeneous - marine origin, in limestone or dolomitic limestone rocks, all formed during the Jurassic. The use of specific fossils for the identification of the cherts is also difficult since these are often not well preserved enough to allow the identification of species that may connect a group of cherts. The size, frequency, and preservation state of the fossils seem to be, then, one of the defining criteria for discerning cherts from different formations, and thus, different geographic areas. These characteristics seem to be observable macroscopically, as well, allowing the cherts from the three different areas and formations – West (including the Lower Jurassic and Upper Jurassic formation cherts of western Algarve), Jordana, and Malhão - to be differentiated without the need for thin sections (@fig-frames-macrocomparison). An exception might be the distinction between the cherts from the West and Malhão - the reddish cherts from Malhão are visually indistinguishable from those from the West with a higher fossil content. The grey cherts of the Malhão Middle Jurassic formation do show a higher concentration of visible round fossils, however the distinction is only possible seen under the stereomicroscope and on a fresh surface, which might hamper the classification of archaeological materials. These distinctions are especially important for archaeological collections, especially those which may be small, with small artifacts, or for the study of older collections to which other (destructive) means of analysis may not be applied.

@@

@@

### 2025-05-03 article2.qmd
More experimentation on aesthetics
Mac work session. Appendix finished. All refs to OSF repos completed.

```{r packages}
#| label: callpackages

@@

```{r dbsetup}
#| label: db-setup

@@

# Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

@@

## Supplementary Information {.unnumbered}

**Online Resource 1**: Data dictionaries for macroscopic (group and individual) and petrographic analyses.

**Online Resource 2**: Terrace and shelter lithotypes complete macroscopic and petrographic description.

**Online resource 3**: Summarised type macroscopic and petrographic features and epoch/stage identification.

>>>>

```{r}
#| label: callpackages

@@

```{r}
#| label: db-setup

@@

# Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia {#sec-chapter3}

@@
